---
title: "Cod diet: historical feeding ratios of cod"
author: "Max Lindmark & Michele Casini"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
  df_print: paged
pdf_document: default
editor_options: 
  chunk_output_type: console
---
  
```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.asp = 0.618,
  fig.align ='center'
)
```

# Intro
Fit gamma hurdle models to stomach content (non-negative, zeroes, continuous) data using `brms` with random ices rectangle as random factors + time period and other predictor variables

```{r lib, message=FALSE}
# Load libraries, install if needed
library(tidyverse); theme_set(theme_classic(base_size = 10))
library(readxl)
library(tidylog)
library(RCurl)
library(viridis)
library(RColorBrewer)
library(patchwork)
library(brms)
library(bayesplot)
library(tidybayes)

# To load entire cache in interactive r session, do: qwraps2::lazyload_cache_dir(path = "R/analysis/cod_feeding_historical_cache/html")

# For parallel processing
options(mc.cores = parallel::detectCores()) 
```

## Define plot theme

```{r read coastline data, message=FALSE, warning=FALSE}
# Define plotting theme for main plot
theme_plot <- function(base_size = 10, base_family = "") {
  theme_light(base_size = 10, base_family = "") +
    theme(
      axis.text.x = element_text(angle = 90),
      axis.text = element_text(size = 8),
      legend.text = element_text(size = 8),
      legend.title = element_text(size = 8),
      legend.position = "bottom",
      legend.key.height = unit(0.2, "cm"),
      legend.margin = margin(0, 0, 0, 0),
      legend.box.margin = margin(-5, -5, -5, -5),
      strip.text = element_text(size = 8, colour = 'black', margin = margin()),
      strip.background = element_rect(fill = "grey90")
      )
}
```

## Read and plot data

```{r read and plot data}
d <- readr::read_csv("/Users/maxlindmark/Desktop/R_STUDIO_PROJECTS/cod_interactions/data/cod_diet_analysis.csv") %>% dplyr::select(-X1)

d <- d %>%
  mutate(ices_rect = as.factor(ices_rect),
         year = as.factor(year),
         time_period = as.factor(time_period),
         quarter = as.factor(quarter),
         depth2_sc = depth - mean(depth)) 

p1 <- ggplot(filter(d, quarter == 1), aes(time_period, FR_tot, color = time_period, fill = time_period)) + 
  ggdist::stat_halfeye(adjust = .5, width = 0.8, .width = 0, justification = -.1, point_colour = NA, alpha = 0.5) + 
  geom_point(shape = 21, color = "white", size = 1.3, alpha = .2, position = position_jitter(seed = 1, width = 0.1)) + 
  geom_boxplot(width = .15, outlier.shape = NA, fill = "white", alpha = 0.2) +  
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  guides(color = FALSE, fill = FALSE) +
  labs(x = "", y = "") +
  coord_flip(ylim = c(0, 0.2)) + # CROPPING FOR BETTER VIEW
  geom_text(data = d %>% filter(quarter == 1) %>% group_by(time_period) %>% summarise(n = n()),
            aes(y = 0.15, x = time_period, label = glue::glue("n = {n}")), nudge_x = 0.3) +
  theme(legend.position = "bottom") + 
  theme_classic(base_size = 10) +
  ggtitle("Quarter 1") +
  NULL

p2 <- ggplot(filter(d, quarter == 4), aes(time_period, FR_tot, color = time_period, fill = time_period)) + 
  ggdist::stat_halfeye(adjust = .5, width = 0.8, .width = 0, justification = -.1, point_colour = NA, alpha = 0.5) + 
  geom_point(shape = 21, color = "white", size = 1.3, alpha = .2, position = position_jitter(seed = 1, width = 0.1)) + 
  geom_boxplot(width = .15, outlier.shape = NA, fill = "white", alpha = 0.2) +  
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  guides(color = FALSE, fill = FALSE) +
  labs(x = "", y = "Feeding ratio") +
  coord_flip(ylim = c(0, 0.2)) + # CROPPING FOR BETTER VIEW
  geom_text(data = d %>% filter(quarter == 4) %>% group_by(time_period) %>% summarise(n = n()),
            aes(y = 0.15, x = time_period, label = glue::glue("n = {n}")), nudge_x = 0.3) +
  theme(legend.position = "bottom") + 
  theme_classic(base_size = 10) +
  ggtitle("Quarter 4") +
  NULL

p1 / p2

# Now do by year
p3 <- ggplot(filter(d, quarter == 1), aes(year, FR_tot, color = time_period, fill = time_period)) + 
  ggdist::stat_halfeye(adjust = .5, width = 0.8, .width = 0, justification = -.1, point_colour = NA, alpha = 0.5) + 
  geom_point(shape = 21, color = "white", size = 1.3, alpha = .2, position = position_jitter(seed = 1, width = 0.1)) + 
  geom_boxplot(width = .15, outlier.shape = NA, fill = "white", alpha = 0.2) +
  facet_grid(~ time_period) + 
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  guides(color = FALSE, fill = FALSE) +
  labs(x = "", y = "") +
  coord_flip(ylim = c(0, 0.2)) + # CROPPING FOR BETTER VIEW
  geom_text(data = d %>% filter(quarter == 1) %>% group_by(time_period, year) %>% summarise(n = n()),
            aes(y = 0.25, x = year, label = glue::glue("n = {n}")), nudge_x = 0.3) +
  theme(legend.position = "bottom") + 
  theme_classic(base_size = 10) +
  ggtitle("Quarter 1") +
  NULL

p3

p4 <- ggplot(filter(d, quarter == 4), aes(year, FR_tot, color = time_period, fill = time_period)) + 
  ggdist::stat_halfeye(adjust = .5, width = 0.8, .width = 0, justification = -.1, point_colour = NA, alpha = 0.5) + 
  geom_point(shape = 21, color = "white", size = 1.3, alpha = .2, position = position_jitter(seed = 1, width = 0.1)) + 
  geom_boxplot(width = .15, outlier.shape = NA, fill = "white", alpha = 0.2) +  
  facet_wrap(~ time_period) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  guides(color = FALSE, fill = FALSE) +
  labs(x = "", y = "Feeding ratio") +
  coord_flip(ylim = c(0, 0.2)) + # CROPPING FOR BETTER VIEW
  geom_text(data = d %>% filter(quarter == 4) %>% group_by(time_period) %>% summarise(n = n()),
            aes(y = 0.25, x = year, label = glue::glue("n = {n}")), nudge_x = 0.3) +
  theme(legend.position = "bottom") + 
  theme_classic(base_size = 10) +
  ggtitle("Quarter 4") +
  NULL

# Filter quarter 1 for this analysis because it has all time periods
d <- d %>% filter(quarter == 1)
```

```{r simpler gamma models, cache=TRUE}
# Read about gamma hurdle models
# https://stats.stackexchange.com/questions/470464/interpreting-residual-plots-for-zero-inflated-linear-mixed-model?rq=1
# https://stats.stackexchange.com/questions/187824/how-to-model-non-negative-zero-inflated-continuous-data
# https://groups.google.com/g/brms-users/c/d5O73pcfaag
# https://discourse.mc-stan.org/t/need-help-understanding-hurdle-hurdle-gamma-models-using-brms/16736
# https://stats.stackexchange.com/questions/477803/need-help-understanding-hurdle-model-specification-and-results-interpretation
# https://seananderson.ca/2014/05/18/gamma-hurdle/
# https://discourse.mc-stan.org/t/estimates-in-hurdle-gamma-model/18934
# https://discourse.mc-stan.org/t/levels-within-levels/8814/4

# Problem that the "top" random effect has few levels?
# https://stats.stackexchange.com/questions/281159/minimum-number-of-repeated-measures-and-levels-per-nested-random-effect
# OK to have time period both as a fixed and in the random part?
# https://stats.stackexchange.com/questions/79360/mixed-effects-model-with-nesting
```

## Fit gamma hurdle model to total prey weight with random `year` effect to each time_period and random `ices_rect` 

```{r random year and rectangle effects, cache=TRUE}
m <- brm(FR_tot ~ 0 + pred_weight_g + depth2_sc + time_period + (0 + time_period|year) + (0 + time_period|ices_rect),
           data = d, family = hurdle_gamma(link = "log", link_shape = "log", link_hu = "logit"),
           chains = 4, iter = 4000, cores = 4, inits = "0",
           control = list(adapt_delta = 0.99, max_treedepth = 15))

summary(m)
plot(m)
```

Inspect model m

```{r inspect m}
# PP check
p1 <- pp_check(m) +
  theme_plot() +
  theme(legend.position = c(0.9, 0.9)) +
  coord_cartesian(xlim = c(0, 0.15)) +
  scale_color_brewer(palette = "Dark2", name = "") +
  NULL

color_scheme_set("gray")
yrep_hu_gamma <- posterior_predict(m, draws = 500)
y <- d$FR_tot

p2 <- ppc_stat(y, yrep_hu_gamma, stat = "median") + theme_plot()
p3 <- ppc_stat(y, yrep_hu_gamma, stat = "mean") + theme_plot()

prop_zero <- function(x) mean(x == 0)
prop_zero(y)

p4 <- ppc_stat(y, yrep_hu_gamma, stat = "prop_zero", binwidth = 0.005)  + theme_plot()

p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

# # Chain convergence
posterior <- as.array(m)
dimnames(posterior)
pal_diag <- rev(brewer.pal(n = 4, name = "Dark2"))

# Chain convergence
p5 <- mcmc_trace(posterior, pars = c("b_time_period1963M1978", "b_time_period1986M1990",
                                     "b_time_period2006M2020", "shape"),
                 facet_args = list(ncol = 2, strip.position = "left")) +
  theme(text = element_text(size = 12), strip.text = element_text(size = 3),
        legend.position = "top") +
  scale_color_manual(values = alpha(pal_diag, alpha = 0.8)) +
  theme_plot()

# Residuals vs fitted
p6 <- d %>%
  add_residual_draws(m) %>%
  ggplot(aes(x = .row, y = .residual)) +
  stat_pointinterval(alpha = 0.5, size = 0.7) +
  theme_plot()

# Quantile dot-plot
get_variables(m)

p7 <- m %>%
  gather_draws(b_time_period1963M1978, b_time_period1986M1990, b_time_period2006M2020) %>%
  mutate(time_period = "1963-1978",
         time_period = ifelse(.variable == "b_time_period1986M1990", "1986-1990", time_period),
         time_period = ifelse(.variable == "b_time_period2006M2020", "2006-2020", time_period)) %>%
  ggplot(aes(x = .value, y = time_period, color = time_period, fill = time_period)) +
  stat_halfeye(alpha = 0.5, color = "grey30", size = 15) +
  labs(y = "") +
  guides(color = FALSE, fill = FALSE) +
  scale_fill_brewer(palette = "Dark2") +
  coord_cartesian(xlim = c(-8, -2)) +
  theme_plot() +
  NULL

mean(d$FR_tot)

p5 / (p6 | p7)

ggsave("figures/FR.png", width = 6.5, height = 6.5, dpi = 600)
```

## Plot random year effects

```{r plot random effects}
m %>%
  spread_draws(b_time_period1963M1978, b_time_period1986M1990, b_time_period2006M2020,
               r_year[year,]) %>%
  mutate(year_mean = ifelse(year < 1979,
                            b_time_period1963M1978 - r_year,
                            b_time_period1986M1990 - r_year),
         year_mean = ifelse(year > 2005,
                            b_time_period2006M2020 - r_year,
                            year_mean)) %>%
  mutate(time_period = ifelse(year < 1979, "1963-1978", "1986-1990"),
         time_period = ifelse(year > 2005, "2006-2020", time_period)) %>% 
  ggplot(aes(y = factor(year), x = year_mean, fill = time_period)) +
  coord_cartesian(xlim = c(-7, -3)) +
  scale_fill_brewer(palette = "Dark2") +
  stat_halfeye(alpha = 0.5)
```

## Fit gamma hurdle model to total prey weight with random `year` effect to each time_period and random `ices_rect` 

```{r random year and rectangle effects, cache=TRUE}
m <- brm(FR_tot ~ 0 + pred_weight_g + depth2_sc + time_period + (0 + time_period|year) + (0 + time_period|ices_rect),
           data = d, family = hurdle_gamma(link = "log", link_shape = "log", link_hu = "logit"),
           chains = 4, iter = 4000, cores = 4, inits = "0",
           control = list(adapt_delta = 0.99, max_treedepth = 15))

summary(m)
plot(m)
```

Inspect model m

## Exit

```{r}
knitr::knit_exit()
```

## Fitting a simple gamma model

```{r fit simple gamma model}
m00 <- brm(FR_tot ~ -1 + pred_weight_g + depth2_sc + time_period,
           data = d, family = hurdle_gamma(link = "log", link_shape = "log", link_hu = "logit"),
           chains = 2, iter = 2000, cores = 2, inits = "0",
           control = list(adapt_delta = 0.9))

summary(m00)
plot(m00)
```

Inspect model m00

```{r inspect m00}
# PP check
p1 <- pp_check(m00) +
  theme_plot() +
  theme(legend.position = c(0.9, 0.9)) +
  coord_cartesian(xlim = c(0, 0.15)) +
  scale_color_brewer(palette = "Dark2", name = "") +
  NULL

color_scheme_set("gray")
yrep_hu_gamma <- posterior_predict(m00, draws = 500)
y <- d$FR_tot

p2 <- ppc_stat(y, yrep_hu_gamma, stat = "median") + theme_plot()
p3 <- ppc_stat(y, yrep_hu_gamma, stat = "mean") + theme_plot()

prop_zero <- function(x) mean(x == 0)
prop_zero(y)

p4 <- ppc_stat(y, yrep_hu_gamma, stat = "prop_zero", binwidth = 0.005)  + theme_plot()

p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

# Chain convergence
posterior <- as.array(m00)
dimnames(posterior)
pal_diag <- rev(brewer.pal(n = 3, name = "Dark2"))

# Chain convergence
p5 <- mcmc_trace(posterior, pars = c("b_time_period1963M1978", "b_time_period1986M1990",
                                     "b_time_period2006M2020", "shape"),
                 facet_args = list(ncol = 2, strip.position = "left")) +
  theme(text = element_text(size = 12), strip.text = element_text(size = 3),
        legend.position = "top") +
  scale_color_manual(values = alpha(pal_diag, alpha = 0.8)) +
  theme_plot()

# Residuals vs fitted
p6 <- d %>%
  add_residual_draws(m00) %>%
  ggplot(aes(x = .row, y = .residual)) +
  stat_pointinterval(alpha = 0.5, size = 0.7) +
  theme_plot()

# Quantile dot-plot
p7 <- m00 %>%
  gather_draws(b_time_period1963M1978, b_time_period1986M1990, b_time_period2006M2020) %>%
  mutate(time_period = "1963-1978",
         time_period = ifelse(.variable == "b_time_period1986M1990", "1986-1990", time_period),
         time_period = ifelse(.variable == "b_time_period2006M2020", "2006-2020", time_period)) %>%
  ggplot(aes(x = .value, y = time_period, color = time_period, fill = time_period)) +
  stat_halfeye(alpha = 0.5, color = "grey30", size = 12) +
  labs(y = "") +
  guides(color = FALSE, fill = FALSE) +
  scale_fill_brewer(palette = "Dark2") +
  theme_plot() +
  NULL

p5 / (p6 | p7)
```

## Add random `year` effect to each time_period

```{r add year random effect, cache=TRUE}
m01 <- brm(FR_tot ~ 0 + pred_weight_g + depth2_sc + time_period + (0 + time_period|year),
           data = d, family = hurdle_gamma(link = "log", link_shape = "log", link_hu = "logit"),
           chains = 2, iter = 2000, cores = 2, inits = "0",
           control = list(adapt_delta = 0.99, max_treedepth = 15))

summary(m01)
plot(m01)
```

Inspect model m01

```{r inspect m01}
# PP check
p1 <- pp_check(m01) +
  theme_plot() +
  theme(legend.position = c(0.9, 0.9)) +
  coord_cartesian(xlim = c(0, 0.15)) +
  scale_color_brewer(palette = "Dark2", name = "") +
  NULL

color_scheme_set("gray")
yrep_hu_gamma <- posterior_predict(m01, draws = 500)
y <- d$FR_tot

p2 <- ppc_stat(y, yrep_hu_gamma, stat = "median") + theme_plot()
p3 <- ppc_stat(y, yrep_hu_gamma, stat = "mean") + theme_plot()

prop_zero <- function(x) mean(x == 0)
prop_zero(y)

p4 <- ppc_stat(y, yrep_hu_gamma, stat = "prop_zero", binwidth = 0.005)  + theme_plot()

p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

# Chain convergence
posterior <- as.array(m01)
dimnames(posterior)
pal_diag <- rev(brewer.pal(n = 3, name = "Dark2"))

# Chain convergence
p5 <- mcmc_trace(posterior, pars = c("b_time_period1963M1978", "b_time_period1986M1990",
                                     "b_time_period2006M2020", "shape"),
                 facet_args = list(ncol = 2, strip.position = "left")) +
  theme(text = element_text(size = 12), strip.text = element_text(size = 3),
        legend.position = "top") +
  scale_color_manual(values = alpha(pal_diag, alpha = 0.8)) +
  theme_plot()

# Residuals vs fitted
p6 <- d %>%
  add_residual_draws(m01) %>%
  ggplot(aes(x = .row, y = .residual)) +
  stat_pointinterval(alpha = 0.5, size = 0.7) +
  theme_plot()

# Quantile dot-plot
get_variables(m01)

p7 <- m01 %>%
  gather_draws(b_time_period1963M1978, b_time_period1986M1990, b_time_period2006M2020) %>%
  mutate(time_period = "1963-1978",
         time_period = ifelse(.variable == "b_time_period1986M1990", "1986-1990", time_period),
         time_period = ifelse(.variable == "b_time_period2006M2020", "2006-2020", time_period)) %>%
  ggplot(aes(x = .value, y = time_period, color = time_period, fill = time_period)) +
  stat_halfeye(alpha = 0.5, color = "grey30", size = 12) +
  labs(y = "") +
  guides(color = FALSE, fill = FALSE) +
  scale_fill_brewer(palette = "Dark2") +
  theme_plot() +
  NULL

p5 / (p6 | p7)
```


## Extract this equation
Doesn't work with any model type that supports gamma, so this is for the random effects only
```{r message=FALSE, warning=FALSE, results="hide"}
# eq <- lme4::lmer(FR_tot ~ 0 + pred_weight_g + depth2_sc + time_period + (0 + time_period|year) + (0 + time_period|ices_rect),
#                  data = d)
# 
# extract_eq(eq)
```

$$
\begin{aligned}
  \operatorname{FR\_tot}_{i}  &\sim N \left(\beta_{0}(\operatorname{pred\_weight\_g}) + \beta_{1}(\operatorname{depth2\_sc}), \sigma^2 \right) \\    
\left(
  \begin{array}{c} 
    \begin{aligned}
      &\gamma_{1j} \\
      &\gamma_{2j} \\
      &\gamma_{3j}
    \end{aligned}
  \end{array}
\right)
  &\sim N \left(
\left(
  \begin{array}{c} 
    \begin{aligned}
      &\mu_{\gamma_{1j}} \\
      &\mu_{\gamma_{2j}} \\
      &\mu_{\gamma_{3j}}
    \end{aligned}
  \end{array}
\right)
, 
\left(
  \begin{array}{ccc}
     \sigma^2_{\gamma_{1j}} & \rho_{\gamma_{1j}\gamma_{2j}} & \rho_{\gamma_{1j}\gamma_{3j}} \\ 
     \rho_{\gamma_{2j}\gamma_{1j}} & \sigma^2_{\gamma_{2j}} & \rho_{\gamma_{2j}\gamma_{3j}} \\ 
     \rho_{\gamma_{3j}\gamma_{1j}} & \rho_{\gamma_{3j}\gamma_{2j}} & \sigma^2_{\gamma_{3j}}
  \end{array}
\right)
 \right)
    \text{, for year j = 1,} \dots \text{,J} \\    
\left(
  \begin{array}{c} 
    \begin{aligned}
      &\gamma_{1k} \\
      &\gamma_{2k} \\
      &\gamma_{3k}
    \end{aligned}
  \end{array}
\right)
  &\sim N \left(
\left(
  \begin{array}{c} 
    \begin{aligned}
      &\mu_{\gamma_{1k}} \\
      &\mu_{\gamma_{2k}} \\
      &\mu_{\gamma_{3k}}
    \end{aligned}
  \end{array}
\right)
, 
\left(
  \begin{array}{ccc}
     \sigma^2_{\gamma_{1k}} & \rho_{\gamma_{1k}\gamma_{2k}} & \rho_{\gamma_{1k}\gamma_{3k}} \\ 
     \rho_{\gamma_{2k}\gamma_{1k}} & \sigma^2_{\gamma_{2k}} & \rho_{\gamma_{2k}\gamma_{3k}} \\ 
     \rho_{\gamma_{3k}\gamma_{1k}} & \rho_{\gamma_{3k}\gamma_{2k}} & \sigma^2_{\gamma_{3k}}
  \end{array}
\right)
 \right)
    \text{, for ices_rect k = 1,} \dots \text{,K}
\end{aligned}
$$

